[Re] Predicting metabolic biomarkers of human inborn errors of metabolism

### 2018-01-02 article/Mondeel_Ogundipe_Westerhoff-2017.tex
[Text] further updates following Aaron's comments
- Move definitions to the introduction
- Refer to the subpanels A and B of Figure 1
- update table IEM names to match original
- up-to-date pdf

The source code is not publicly available and the computation
environment used was not mentioned in the publication. The
\href{https://github.com/opencobra/cobratoolbox}{MATLAB COBRA toolbox}
contains code to reproduce the analysis by Thiele et al. but this was
also performed in MATLAB.

Here we here present an implementation of the biomarker prediction
method originally proposed by Shlomi et al., programmed in Python, and
reproduce the results presented in Figure 1 and 2 of the original
publication. These figures concern the method's application to a simple
metabolic map and then to a set of \(17\) amino acid metabolism
disorders. Our Python module provides the biomarker prediction algorithm
and is agnostic of the map used. It can therefore be applied to other
maps than the human metabolic reconstruction considered by Shlomi et al.
Jupyter notebooks are provided that generate the illustrative map and
reproduce all results referred to in Figures 1 and 2 of Shlomi et al.
Our implementation is based on information given in the original paper.
However, some details critical to the reproduction were left unexplained
there and are highlighted here. We also include a more detailed
discussion of one of the amino acid IEMs to examine the robustness of
the predictions made using this approach.

@@

Flux ranges for boundary metabolites are calculated by applying flux
variability analysis \autocite{Mahadevan2003}. Mathematically, flux
variability analysis for a specific exchange reaction \(v_i\) entails
the following linear programming problem:

\begin{align}
\text{Maximize } &\pm v_i, \text{ such that for all k} \nonumber \\
S v &= 0 \nonumber \\
Z  &\geq \phi Z_{\text{opt}} \nonumber \\
\alpha_k \leq & ~ v_k \leq \beta_k 
\end{align}

Here \(v\) is the column vector of fluxes representing all reactions in
the model, \(v_i\) is the flux of the exchange reaction of a given
boundary metabolite \(i\), \(\alpha_k\) is the, possibly negative, lower
bound for reaction \(k\) and similarly \(\beta_k\) is the, possibly
negative, upper bound for reaction \(k\). \(Z = c^T ~ v\) is the
objective function defined for the map, entailing a linear combination,
defined by column vector c, of one or more fluxes, and
\(Z_{\text{max}}\) its maximal value. We define \(0 \leq \phi \leq 1\)
to be an arbitrary minimal fraction of the maximal objective function
value that has to be achieved. When \(\phi=1\) we ask for the range of
flux allowed for reaction \(v_i\) while maintaining the maximal value of
the linear combination of objective fluxes. When setting \(\phi = 0\),
one is asking for the allowable flux range through reaction \(v_i\)
regardless of the value of the objective function (although the sign has
to be maintained). The latter, i.e. \(\phi=0\), is the case considered
in the method by Shlomi et al.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Compute the exchange flux interval of \(m\), when \(r\) is forced to
  be active with a flux of at least \(\epsilon \geq 0\). By default
  \(\epsilon = 1\). We refer to this as the forward wild-type interval
  \(WT_{r,m}^+\).
\item
  For reversible reactions \(r\), if the forward interval yields a flux
  range of \([0,0]\), constrain the flux to be negative and lower than
  \(-\epsilon \leq 0\), and compute the backward wild-type interval
  \(WT_{r,m}^-\).
\item
  The wild-type interval \(WT_{r,m}\) is equal to the union of the
  forward and backward intervals.\\
\item
  Compute the exchange flux interval of \(m\) when \(r\) is forced to be
  inactive by temporarily setting the lower and upper bound of \(r\) to
  zero. This interval is denoted by \(M_{r,m}\).
\item
  Compare the wild-type and mutant intervals and predict whether the
  external concentration of \(m\) increases, decreases or remains
  unchanged when reaction \(r\) is forced to be inactive. This is done
  by the following rule: if both boundaries are higher in the mutant
  simulation than the corresponding boundaries in the wild-type
  simulation, then \(m\) is predicted to be a biomarker and elevated. If
  both boundaries are lower in the mutant simulation as compared to the
  wild-type simulation, then the metabolite is predicted to be a
  biomarker and reduced. If the predicted changes are smaller than 10\%,
  \(m\) is a low confidence biomarker. When the wild-type and mutant
  intervals are disjoint intervals, \(m\) is a highly confident (H.C)
  biomarker.
\end{enumerate}

@@

@@

\subsection{\texorpdfstring{Reproduction of Figure 1 in
\autocite{Shlomi2009}}{Reproduction of Figure 1 in {[}@Shlomi2009{]}}}\label{reproduction-of-figure-1-in-shlomi2009}

@@

The original publication's Figure 1 exemplified the biomarker prediction
method for an illustrative network. It is reproduced here as Figure
~\ref{fig:Shlomi2009_figure_1}. The simple nature of the example serves
to explain the reasoning behind the method and also helped us validate
our implementation of the method. This repository provides an SBML file
of the network, the Jupyter notebook that generates it and a Jupyter
notebook reproducing the biomarker prediction.

The network in Figure ~\ref{fig:Shlomi2009_figure_1} consists of seven
metabolites, \(6\) of which (i.e.~all except M3) have exchange
reactions. We consider a hypothetical deletion of the enzyme catalyzing
the conversion of M1 into M2 coupled to the conversion of M3 into M4.
Shlomi et al. graphically provide the biomarker prediction for this
network in their Figure 1. Here, we do so numerically and this paper is
accompanied by a Jupyter notebook reproducing the biomarker prediction.
In our analysis, all exchange reactions were given inward bounds of
\(-10\) and outward bounds of \(+1000\) so that the internal reaction
bounds (of \(\pm 1000\)) are never reached. Our Table
~\ref{tbl:toy_model_results} shows that all exchangeable metabolites
except M5 and M7 are predicted to be robust biomarkers.

@@

\hypertarget{tbl:amino_acid_iems}{}
\begin{longtable}[]{@{}llll@{}}
\caption{\label{tbl:amino_acid_iems}We used our implementation to
reproduce the biomarkers for the list of inborn errors of metabolism
shown in Figure 2 in \autocite{Shlomi2009}. Using the supplementary
Excel table of \autocite{Shlomi2009} linking IEMs to causative genes and
affected reactions, we filtered out those that are listed here. The only
considered biomarkers are the \(20\) proteinogenic amino acids. We
report all biomarkers the change of which exceeds the 10\% threshold and
categorize them by their qualitative prediction: elevated or reduced
serum levels. }\tabularnewline
\toprule
IEM & Affected reactions & Elevated & Reduced\tabularnewline
\midrule
\endfirsthead
\toprule
IEM & Affected reactions & Elevated & Reduced\tabularnewline
\midrule
\endhead
S-Adenosylhomocysteine & SEAHCYSHYD, AHCi & L-Methionine &
L-Cysteine\tabularnewline
hydrolase & & &\tabularnewline
Alkaptonuria & HGNTOR & L-Tyrosine &\tabularnewline
Arginase deficiency & ARGN & &\tabularnewline
Cystinuria & CYSTSERex, & &\tabularnewline
& SERLYSNaex & &\tabularnewline
Lysinuric protein & SERLYSNaex & &\tabularnewline
intolerance & & &\tabularnewline
Glutamate formimino- & FTCD, GluForTx & L-Histidine &\tabularnewline
transferase deficiency & & &\tabularnewline
Histidinemia & HISDr & L-Histidine &\tabularnewline
Homocystinuria & CYSTS, MTHFR3, & L-Methionine &
L-Cysteine\tabularnewline
& METS & &\tabularnewline
Hyperprolinemia type I & PRO1xm, PROD2m & &\tabularnewline
Maple syrup & OIVD1m, & L-Valine, &\tabularnewline
urine disease & OIVD2m, & L-Isoleucine &\tabularnewline
& OIVD3m & L-Leucine, &\tabularnewline
Methionine adenosyl- & SELMETAT, METAT & L-Methionine &
L-Cysteine\tabularnewline
transferase deficiency & & &\tabularnewline
methylmalonic acidemia & CBLATm, CBL2tm, & L-Isoleucine &\tabularnewline
& MMMm & &\tabularnewline
Phenylketonuria & PHETHPTOX2 & L-Phenylalanine &
L-Tyrosine\tabularnewline
Tetrahydrobiopterin & DHPR & &\tabularnewline
deficiency & & &\tabularnewline
Tyrosinemia type 1 & FUMAC & L-Tyrosine &\tabularnewline
Tyrosinemia type III & PPOR, 34HPPOR & L-Tyrosine &\tabularnewline
Glycine encephalopathy & GCC2am, GCC2bim, & &\tabularnewline
& GCC2cm, GCCam, & &\tabularnewline
& GCCbim, GCCcm & &\tabularnewline
\bottomrule
\end{longtable}

>>>>

Below, we first introduce the essential vocabulary and concepts
necessary for understanding the method proposed by Shlomi et al.

@@

Flux ranges for boundary metabolites, consistent with the requirements
of a steady state and constraints on reaction reversibilities, may be
calculated by applying flux variability analysis
\autocite{Mahadevan2003}. Mathematically, flux variability analysis for
a specific exchange reaction \(v_i\) entails the following linear
programming problem:

\begin{align}
\text{Maximize } &\pm v_i, \text{ such that for all k} \nonumber \\
S v &= 0 \nonumber \\
Z  &\geq \phi Z_{\text{max}} \nonumber \\
\alpha_k \leq & ~ v_k \leq \beta_k 
\end{align}

Here \(v\) is the column vector of fluxes representing all reactions in
the model, \(v_i\) is the flux of the exchange reaction of a given
boundary metabolite \(i\), \(\alpha_k\) is the, possibly negative, lower
bound for reaction \(k\) and similarly \(\beta_k\) is the, possibly
negative, upper bound for reaction \(k\). The reaction bounds are the
\(V_{max}\)'s of the reactions and are part of the metabolic network
definition, typically they are set to either \(0\) or \(\pm 1000\) if
the true, biological, \(V_{max}\)'s are not known. These values allow
specification of irreversible reactions by setting the lower bound (or
upper bound) to zero. The index \(i\) specifies a single reaction that
is to be maximized, whereas the index \(k\) is used to index the bound
constraints that exist for each reaction, including \(v_i\). \(v_i\) may
be maximized in the forward (positive flux) or the reverse (negative
flux) direction if allowed by the bounds on \(v_i\). This is indicated
by the \(\pm\) symbol. \(S\) is the stoichiometry matrix for the
metabolic network of size \(mxr\) where \(m\) is the number of
metabolites and \(r\) the number of reactions. The numeric, typically
integer, elements of \(S\) represent the stoichiometry coefficients of
each metabolite \(m\) in each reaction \(r\). Consequently, for large
networks, this matrix is sparse (has many zero entries). \(Z = c^T ~ v\)
is the objective function defined for the map, entailing a linear
combination, defined by column vector \(c\), of one or more reactions.
The vector \(c\) is an indicator of objective reactions, i.e.~it
contains a value of \(1\) in rows corresponding to fluxes that are to be
included in the objective, and zeros in all other rows. Due to the
vector multiplication \(c^T ~ v\), the objective \(Z\) sums the fluxes
of the reactions that correspond to rows containing a value of \(1\) in
\(c\). \(Z_{\text{max}}\) denotes the maximal value of the objective
function. We define \(0 \leq \phi \leq 1\) to be an arbitrary minimal
fraction of the maximal objective function value that has to be
achieved. When \(\phi=1\) we ask for the range of flux allowed for
reaction \(v_i\) while maintaining the maximal value of the linear
combination of objective fluxes. When setting \(\phi = 0\), one is
asking for the allowable flux range through reaction \(v_i\) regardless
of the value of the objective function (although the sign has to be
maintained). The latter, i.e. \(\phi=0\), is the case considered in the
method by Shlomi et al.

\subsection{Reference implementation}\label{reference-implementation}

The source code for the original publication is not publicly available
and the computation environment used was not mentioned in the
publication. The \href{https://github.com/opencobra/cobratoolbox}{MATLAB
COBRA toolbox} contains code to reproduce the analysis by Thiele et al.
but this was also performed in MATLAB.

We here present an implementation of the biomarker prediction method
originally proposed by Shlomi et al., programmed in Python, and
reproduce the results presented in Figure 1A-B and 2 of the original
publication. These figures concern the method's application to a simple
metabolic map and then to a set of \(17\) amino acid metabolism
disorders. Our Python module provides the biomarker prediction algorithm
and is agnostic of the map used. It can therefore be applied to other
maps than the human metabolic reconstruction considered by Shlomi et al.
Jupyter notebooks are provided that generate the illustrative map and
reproduce all results referred to in Figures 1A-B and 2 of Shlomi et al.
Our implementation is based on information given in the original paper.
However, some details critical to the reproduction were left unexplained
there and are highlighted here. We also include a more detailed
discussion of one of the amino acid IEMs to examine the robustness of
the predictions made using this approach. We feel this adds to the
utility of this reference implementation and the understanding of the
approach introduced by Shlomi et al.

\subsection{Implementation of the
algorithm}\label{implementation-of-the-algorithm}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Compute the exchange flux interval of \(m\), when \(r\) is forced to
  be active with a flux of at least \(\epsilon \geq 0\). By default
  \(\epsilon = 1\). We refer to this as the forward wild-type interval
  \(WT_{r,m}^+\).
\item
  For reversible reactions \(r\), if the forward interval yields a flux
  range of \([0,0]\), constrain the flux to be negative and lower than
  \(-\epsilon \leq 0\), and compute the backward wild-type interval
  \(WT_{r,m}^-\).
\item
  The wild-type interval \(WT_{r,m}\) is equal to the union of the
  forward and backward intervals.
\item
  Compute the exchange flux interval of \(m\) when \(r\) is forced to be
  inactive by temporarily setting the lower and upper bound of \(r\) to
  zero. This interval is denoted by \(M_{r,m}\).
\item
  Compare the wild-type and mutant intervals and predict whether the
  external concentration of \(m\) increases, decreases or remains
  unchanged when reaction \(r\) is forced to be inactive. This is done
  by the following rule: if both boundaries are higher in the mutant
  simulation than the corresponding boundaries in the wild-type
  simulation, then \(m\) is predicted to be a biomarker and elevated. If
  both boundaries are lower in the mutant simulation as compared to the
  wild-type simulation, then the metabolite is predicted to be a
  biomarker and reduced. If the predicted changes are smaller than 10\%,
  \(m\) is a low confidence biomarker. When the wild-type and mutant
  intervals are disjoint intervals, \(m\) is a highly confident (H.C)
  biomarker.
\end{enumerate}

@@

In this section we start with an analysis of the sensitivity of the
method which was not clearly discussed in the original publication but
is easily facilitated with our reference implementation. Subsequently,
we discuss the reproduction of Figures 1 and 2 from the original
publication.

@@

\subsection{\texorpdfstring{Reproduction of Figure 1 panel A and B in
\autocite{Shlomi2009}}{Reproduction of Figure 1 panel A and B in {[}@Shlomi2009{]}}}\label{reproduction-of-figure-1-panel-a-and-b-in-shlomi2009}

@@

The original publication's Figure 1, panel A and B, exemplified the
biomarker prediction method for an illustrative network. It is
reproduced here as Figure ~\ref{fig:Shlomi2009_figure_1}. The simple
nature of the example serves to explain the reasoning behind the method
and also helped us validate our implementation of the method. This
repository provides an SBML file of the network, the Jupyter notebook
that generates it and a Jupyter notebook reproducing the biomarker
prediction.

The network in Figure ~\ref{fig:Shlomi2009_figure_1} consists of seven
metabolites, \(6\) of which (i.e.~all except M3) have exchange
reactions. We consider a hypothetical deletion of the enzyme catalyzing
the conversion of M1 into M2 coupled to the conversion of M3 into M4.
Shlomi et al. graphically provide the biomarker prediction for this
network in their Figure 1A. Here, we do so numerically and this paper is
accompanied by a Jupyter notebook reproducing the biomarker prediction.
In our analysis, all exchange reactions were given inward bounds of
\(-10\) and outward bounds of \(+1000\) so that the internal reaction
bounds (of \(\pm 1000\)) are never reached. Our Table
~\ref{tbl:toy_model_results} shows that all exchangeable metabolites
except M5 and M7 are predicted to be robust biomarkers.

@@

\hypertarget{tbl:amino_acid_iems}{}
\begin{longtable}[]{@{}llll@{}}
\caption{\label{tbl:amino_acid_iems}We used our implementation to
reproduce the biomarkers for the list of inborn errors of metabolism
shown in Figure 2 in \autocite{Shlomi2009}. Using the supplementary
Excel table of \autocite{Shlomi2009} linking IEMs to causative genes and
affected reactions, we filtered out those that are listed here. The only
considered biomarkers are the \(20\) proteinogenic amino acids. We
report all biomarkers the change of which exceeds the 10\% threshold and
categorize them by their qualitative prediction: elevated or reduced
serum levels. }\tabularnewline
\toprule
IEM & Affected reactions & Elevated & Reduced\tabularnewline
\midrule
\endfirsthead
\toprule
IEM & Affected reactions & Elevated & Reduced\tabularnewline
\midrule
\endhead
S-Adenosylhomocysteine & SEAHCYSHYD, AHCi & L-Methionine &
L-Cysteine\tabularnewline
hydrolase & & &\tabularnewline
Alkaptonuria & HGNTOR & L-Tyrosine &\tabularnewline
Argininemia & ARGN & &\tabularnewline
Cystinuria & CYSTSERex, & &\tabularnewline
& SERLYSNaex & &\tabularnewline
Lysinuric protein & SERLYSNaex & &\tabularnewline
intolerance & & &\tabularnewline
Glutamate formimino- & FTCD, GluForTx & L-Histidine &\tabularnewline
transferase deficiency & & &\tabularnewline
Histidinemia & HISDr & L-Histidine &\tabularnewline
Homocystinuria & CYSTS, MTHFR3, & L-Methionine &
L-Cysteine\tabularnewline
& METS & &\tabularnewline
Hyperprolinemia type I & PRO1xm, PROD2m & &\tabularnewline
Maple syrup & OIVD1m, & L-Valine, &\tabularnewline
urine disease & OIVD2m, & L-Isoleucine &\tabularnewline
& OIVD3m & L-Leucine, &\tabularnewline
Methionine adenosyl- & SELMETAT, METAT & L-Methionine &
L-Cysteine\tabularnewline
transferase deficiency & & &\tabularnewline
methylmalonic acidemia & CBLATm, CBL2tm, & L-Isoleucine &\tabularnewline
& MMMm & &\tabularnewline
Phenylketonuria & PHETHPTOX2 & L-Phenylalanine &
L-Tyrosine\tabularnewline
Phenylketonuria II & DHPR & &\tabularnewline
Tyrosinemia type I & FUMAC & L-Tyrosine &\tabularnewline
Tyrosinemia type III & PPOR, 34HPPOR & L-Tyrosine &\tabularnewline
Glycine encephalopathy & GCC2am, GCC2bim, & &\tabularnewline
& GCC2cm, GCCam, & &\tabularnewline
& GCCbim, GCCcm & &\tabularnewline
\bottomrule
\end{longtable}
